Longitudinal Manifold Estimation

### 2023-02-15 paper/main.tex
add section 2 outline

Because the algorithm proposed in this article relies on the same definitions and notation as those introduced in Meng and Eloyan (2021), we provide it here. To generalize the principal curve framework proposed in Hastie and Stuetzle (1989) to intrinsic dimensions $d > 1$, Meng and Eloyan introduce the function space 
\[%
  C_{\infty}(\mathbb{R}^d \to \mathbb{R}^D) = \left\{f \in C()\right\}
.\]%

@@

>>>>

The notation below is used to build to a definition of principal manifolds [still need to formally write out]:
\begin{itemize}
  \item Observed data $\left\{x_i\right\}_{i=1}^I \subset \mathbb{R}^D$
  \item Associated latent parameterizations $\left\{t_i\right\}_{i=1}^I \subset \mathbb{R}^d$
  \item $C_{\infty}\left(\mathbb{R}^d \to \mathbb{R}^D\right) = \left\{f \in C(\mathbb{R}^d \to \mathbb{R}^D): \lim_{\|t\|_{\mathbb{R}^d} \to \infty} \|f(t)\|_{\mathbb{R}^D} = \infty\right\}$ 
  \item $\nabla f$ denotes the gradient of $f$
  \item $\nabla^{\bigotimes 2}f$ denotes the Hessian matrix of $f$
  \item $\|\nabla^{\bigotimes 2}f\|_{\mathbb{R}^{d \times d}} = \left(\sum_{i, j = 1}^{d} |\frac{\partial^2f}{\partial t_i \partial t_j}(t)|^2\right)^{\frac{1}{2}}$
  \item $\|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l = 1}^{D}\|\nabla^{\bigotimes 2}f_l\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D}\int_{\mathbb{R}^d}\sum_{i,j = 1}^{d}|\frac{\partial^2f_l}{\partial t_i \partial t_j}(t)|^2dt$ represents the total squared curvature of manifold $M_f^{d}$
  \item $\mathcal{K}_{\lambda, \mathbb{P}}(f, g) = \mathbb{E}\|X - f(\pi_{g}(X))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2$, with $\mathcal{K_{\lambda, \mathbb{P}}}(f) = \mathcal{K}_{\lambda, \mathbb{P}}(f, f)$
  \item A manifold $M_{f^*}^{d}$ determined by $f^*$ is a principal manifold for random $D$-vector $X$ with compact support $\text{supp}(\mathbb{P})$ with the tuning parameter $\lambda$ if \[%
    f_{\lambda}^{*} = \arg \min_{f \in \mathcal{F}(\mathbb{P})}\mathcal{K}_{\lambda, \mathbb{P}}(f)
  ,\]%
    where \[%
      \mathcal{F}(\mathbb{P}) = \left\{f \in C_{\infty} \cap \nabla^{-biot 2}L^2(\mathbb{R}^{d} \to \mathbb{R}^{D}): \sup_{x \in \text{supp}(\mathbb{P})}\|\pi_f(x)\|_{\mathbb{R}^{d}} = 1\right\}
    .\]%
  \item $\pi_{f_\lambda^{*}}(X)$ gives a $d$-dimensional parameterization of $X$
\end{itemize}

Given this definition, we now summarize the estimation method provided. This algorithm has three main steps: data reduction, fitting, and tuning. In the reduction step, the HDMDE algorithm is used to summarize the data $\left\{x_i\right\}_{i=1}^{I}$ by points $\left\{\mu_j\right\}_{j=1}^{N}$ with $N \leq I$, and each $\mu_j$ associated with weight $\theta_j$. The fitting stage of the algorithm finds estimated functions 
\[%
  \hat{f}_\lambda = \arg\min_f \mathcal{K}_{\lambda, \hat{Q}_N}(f)
,\]%
where 
\[%
  \mathcal{K}_{\lambda, \hat{Q}_N}(f) = \sum_{j = 1}^{N}\theta_j \|\mu_j - f(\pi_f(\mu_j))\|_{\mathbb{R}^{D}}^2 + \lambda\|\nabla^{\bigotimes 2}f\|_{L^2(\mathbb{R}^{d})}^2
\]% 
approximates $\mathcal{K}_{\lambda, \mathbb{P}}(f)$. The tuning stage chooses $\lambda^{*} = \arg\min_{\lambda > 0}(\mathcal{D}(\hat{f}_\lambda))$, where 
\[%
  \mathcal{D}(\hat{f}_\lambda) = \frac{1}{I}\sum_{i=1}^{I}\|x_i - \hat{f}_\lambda(\pi_{\hat{f}_\lambda}(x_i))\|_{\mathbb{R}^{D}}^2
.\]%

Meng and Eloyan (2021) show that the optimal function will be of spline form. Specific details of the HDMDE algorithm (reduction step) and the PME algorithm (fitting and tuning steps) are given in Meng and Eloyan (2021).

@@

Contributions:

Limitations:
\begin{enumerate}
  \item Self-consistency condition means that the model cannot fit to self-intersecting manifolds. This is why partitioning is necessary.
  \item Partitioning approach is somewhat ad hoc and may not work in every situation. In testing it struggled when the structure shape changed drastically, or when orientation was not equivalent. There is not a clear approach to automatic partitioning.
  \item The process of joining the partitions may introduce error.
  \item Algorithm initialization is sensitive to magnitude of change between time points.
  \item Computation time.
\end{enumerate}

Future Work:
